---
title: "Hurricanes EDA"
output: pdf_document
---

# Data Sources
**amo_by_month.csv**:  
https://climatedataguide.ucar.edu/sites/default/files/amo_monthly.10yrLP.txt  
(Manually generated CSV file from the textual data)  

**category4_atlantic_by_decade.csv**:  
https://en.wikipedia.org/wiki/List_of_Category_4_Atlantic_hurricanes  
(Manually generated CSV file from the tables)  

**category4_atlantic_by_year.csv**:  
https://en.wikipedia.org/wiki/List_of_Category_4_Atlantic_hurricanes  
(Manually generated CSV file from the tables)  

**category5_atlantic_by_decade.csv**:  
https://en.wikipedia.org/wiki/List_of_Category_5_Atlantic_hurricanes  
(Manually generated CSV file from the tables)  

**category5_atlantic_by_year.csv**:  
https://en.wikipedia.org/wiki/List_of_Category_5_Atlantic_hurricanes  
(Manually generated CSV file from the tables)  

**tropical_storm_data**:  

# Plot Bar Plots of Hurricanes by Decade
## Category 4 Hurricanes
```{r}
Cat4.Atlantic.decade <-
  read.csv('datasets/category4_atlantic_by_decade.csv')
Cat4.Atlantic.decade
```

```{r}
barplot(
  Cat4.Atlantic.decade$frequency,
  names.arg = Cat4.Atlantic.decade$decade,
  main = 'Category 4 Hurricanes by Decade',
  xlab = 'Decade',
  ylab = 'Frequency',
  las = 2
)
grid()
```

## Category 5 Hurricanes
```{r}
Cat5.Atlantic.decade <-
  read.csv('datasets/category5_atlantic_by_decade.csv')
Cat5.Atlantic.decade
```

```{r}
barplot(
  Cat5.Atlantic.decade$frequency,
  names.arg = Cat5.Atlantic.decade$decade,
  main = 'Category 5 Hurricanes by Decade',
  xlab = 'Decade',
  ylab = 'Frequency',
  las = 2
)
grid()
```

# Calculate the Highest Frequency of Hurricanes Over the Decades
```{r}
Cat4.max_freq = max(Cat4.Atlantic.decade$frequency)
Cat4.max_freq
```
```{r}
Cat5.max_freq = max(Cat5.Atlantic.decade$frequency)
Cat5.max_freq
```

# Calculate Number of Decades
```{r}
num_decades = length(Cat4.Atlantic.decade$decade)
num_decades
```

# Function to Count the Number of Decades Having the Same Hurricane Frequencies
Example: Number of Decades where there was 1 occurrence of Category 4 Hurricane is 2,  
Number of Decades where there were 2 occurrences of Category 4 Hurricanes is 3, etc.
```{r}
get_freqs <- function(Cat, freqs_list) {
  for (i in Cat$frequency) {
    freqs_list[i + 1] = freqs_list[i + 1] + 1
  }
  return(freqs_list)
}
```

```{r}
Cat4.freqs =  c(0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0)
Cat4.freqs = get_freqs(Cat4.Atlantic.decade, Cat4.freqs)
Cat4.freqs
```
```{r}
Cat5.freqs =  c(0, 0, 0, 0, 0, 0, 0, 0, 0)
Cat5.freqs = get_freqs(Cat5.Atlantic.decade, Cat5.freqs)
Cat5.freqs
```

# Plot Number of Hurricanes in a Decade
```{r}
barplot(
  Cat4.freqs,
  names.arg = 0:Cat4.max_freq,
  main = 'Category 4 Hurricanes by Decade',
  xlab = 'Number of Hurricanes in a Decade',
  ylab = 'Frequency',
  las = 2
)
grid()
```


```{r}
barplot(
  Cat5.freqs,
  names.arg = 0:Cat5.max_freq,
  main = 'Category 5 Hurricanes by Decade',
  xlab = 'Number of Hurricanes in a Decade',
  ylab = 'Frequency',
  las = 2
)
grid()
```

# Calculate lambda Value for Poisson Distribution
```{r}
Cat4.lambda = mean(Cat4.Atlantic.decade$frequency)
Cat4.lambda
```

```{r}
Cat5.lambda = mean(Cat5.Atlantic.decade$frequency)
Cat5.lambda
```

# Plot Poisson Distribution Over Hurricane Frequencies
```{r}
plot(
  0:Cat4.max_freq,
  Cat4.freqs,
  main = 'Poisson Distribution Over Category 4 Hurricane Frequencies',
  xlab = 'Hurricane Frequency in a Decade',
  ylab = 'Count of Decades'
)
lines(0:Cat4.max_freq,
      dpois(0:Cat4.max_freq, Cat4.lambda) * sum(Cat4.freqs),
      lwd = 2)
grid()
```

```{r}
plot(
  0:Cat5.max_freq,
  Cat5.freqs,
  main = 'Poisson Distribution Over Category 5 Hurricane Frequencies',
  xlab = 'Hurricane Frequency in a Decade',
  ylab = 'Count of Decades'
)
lines(0:Cat5.max_freq,
      dpois(0:Cat5.max_freq, Cat5.lambda) * sum(Cat5.freqs),
      lwd = 2)
grid()
```

# Plot Q-Q Plots
```{r}
qqplot(
  dpois(0:Cat4.max_freq, Cat4.lambda) * sum(Cat4.freqs),
  Cat4.freqs,
  main = 'Poisson Q-Q Plot for Category 4 Hurricanes',
  xlab = 'Theoretical Quantiles',
  ylab = 'Sample Quantiles',
  xlim = c(0, 4),
  ylim = c(0, 4)
)
abline(0, 1)
```

```{r}
qqplot(
  dpois(0:Cat5.max_freq, Cat5.lambda) * sum(Cat5.freqs),
  Cat5.freqs,
  main = 'Poisson Q-Q Plot for Category 5 Hurricanes',
  xlab = 'Theoretical Quantiles',
  ylab = 'Sample Quantiles',
  xlim = c(0, 4),
  ylim = c(0, 4)
)
abline(0, 1)
```

# Read Monthly Atlantic Multidecadal Oscillation Index CSV File
```{r, results='hide'}
AMO.month <- read.csv('datasets/amo_by_month.csv')
AMO.month
```

# Remove Outlier Values
```{r, results='hide'}
AMO.month = AMO.month[c(-146:-152), ]
AMO.month
```

# Plot AMO Index (Yearly Mean)
```{r}
AMO.month.means = rowMeans(AMO.month[, -1])
AMO.month.means.ts = ts(AMO.month.means, frequency = 1, start = 1870)
plot(AMO.month.means.ts,
     main = 'AMO Index (Yearly Mean)',
     xlab = 'Year',
     ylab = 'Temperature Anomaly')
abline(h = 0, lwd = 2)
grid()
```
